The Proofs

### 2024-04-03 proofs/main.tex
add proofs: elaboration

The types and terms of \fexists have an additional polarization structure.
Despite that, it is closely related to the well-known \systemf.
To demonstrate that, we provide two translations in both ways:
the polarization from \systemf to \fexists and the depolarization from 
\fexists to \systemf. These translations are done at the level of types and terms,
and the expected typing preservation properties are proved.

\begin{figure}[h]
  \ottdefnSFtyLabeled{}
  \ottdefnSFtyALabeled{}
  \caption{Typing rules of \systemf}
  \label{fig:systemf-typing}
\end{figure}

\begin{figure}[h]
  \ottdefnSFWFLabeled{}
\end{figure}

\begin{figure}[h]
  \ottfundefnunpolP{}
  \ottfundefnunpolN{}
\end{figure}

\begin{figure}[h]
  \ottdefnELNLabeled{}
  \ottdefnELPLabeled{}
  \caption{Subtyping elaboration from \fexists to \systemf}
  \label{fig:subt-elaboration}
\end{figure}

The term-level translation is defined not solely for terms, but for typing
derivations. The reason for that is that the terms of \systemf and \fexists
contain different typing information that cannot be reconstructed without the
appropriate derivation tree. For instance, the lambda expressions in \systemf do
not have the bound variable annotation, while the lambda expressions in \fexists
do; on the other hand, \fexists has rich subtyping, which cannot be expressed in
in terms of polymorphic type instantiation of \systemf.

\begin{figure}[h]
  \ottdefnTELPLabeled{}
  \caption{Typing elaboration from \fexists to \systemf}
  \label{fig:typing-elaboration}
\end{figure}

\begin{figure}[h]
  \ottdefnTELNLabeled{}
\end{figure}

\begin{figure}[h]
  \ottdefnTELAppLabeled{}
\end{figure}

\begin{figure}[h]
  \ottdefnSFfPolLabeled{}
  \caption{Polarization: translation from \systemf to \fexists}
  \label{fig:systemf-polarization}
\end{figure}

% \lemmaDepolarSoundness*

% \lemmaPolSubst

% \lemmaPolRemovable

% \lemmaTypPolWf*

% \lemmaPolarSoundness

@@

>>>>

@@

Although based on \systemf, \fexists has an additional polarization structure.
To demonstrate the relation between these systems we establish translations in
both ways: the polarization from \systemf to \fexists and the depolarization
from \fexists to \systemf. These translations are done at the level of types and
terms, and the expected typing preservation properties are proved.

First, let us agree on the variant of \systemf that we use to establish the
relation with \fexists. At the type level, we have variables, functional arrows,
and universal quantifiers, and define existential quantifiers as a syntactic
sugar using standard encodings. At the term level, we have variables,
unannotated lambda abstractions, type abstractions, and term-level applications,
but not type applications (they can be done implicitly); for convenience, we
introduce multi-argument application, let-blinders, and existential constructors
and eliminators as syntactic sugar (see \cref{fig:systemf-grammar}).

\begin{figure}
  \begin{minipage}[t]{0.39\linewidth}
    \vspace*{0pt}
    \ottgrammartabular{
      \ottfT\ottprodnewline
      \ottprodline{}{}{}{}{}{}\ottprodnewline
      \ottprodline{}{}{}{}{}{}\ottprodnewline
      \ottprodline{|}{[[∃as.fT]]}{}{}{}{}\ottprodnewline
      \ottprodline{}{\equiv [[∀b.(∀as.(fT → b)) → b]]}{}{}{}{}\ottinterrule
    }
  \end{minipage}
  \begin{minipage}[t]{0.6\linewidth}
    \vspace*{0pt}
    \ottgrammartabular{
      \ottft\ottprodnewline
      \ottprodline{}{}{}{}{}{}\ottprodnewline
      \ottprodline{|}{[[ft' ● fargs]]}{\equiv (([[ft']] [[ft1]])\, \dots )\,[[ftn]]}{}{}{}\ottprodnewline
      \ottprodline{|}{[[let x = ft1 ; ft2]]} {\equiv [[(λx.ft2) ft1]]}{}{}{}\ottprodnewline
      \ottprodline{|}{[[pack ft : fT]]}{\equiv [[Λb. λf. f ft]]}{}{}{}\ottprodnewline
      \ottprodline{|}{[[unpack ( as , x ) = ft1 ; ft2]]}{\equiv [[ft1 (Λ as. λx. ft2)]]}{}{}{}\ottinterrule
    }
  \end{minipage}
  \caption{Grammar of \systemf}
  \label{fig:systemf-grammar}
\end{figure}

\begin{figure}
  \ottdefnSFtyLabeledTwoColumns{}
  \caption{Typing rules of \systemf}
  \label{fig:systemf-typing}
\end{figure}

\begin{observation}
  The following rules are admissible in \systemf: \hfill

  \begin{minipage}{0.5\textwidth}
    \ottusedrule{\ottdruleSFtyAUnpackLabeled{}}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \ottusedrule{\ottdruleSFtyAPackLabeled{}}
    \ottusedrule{\ottdruleSFtyALetLabeled{}}
  \end{minipage}
\end{observation}

% \begin{figure}[h]
%   \ottdefnSFWFLabeled{}
% \end{figure}

The translation between the types is defined in 
\cref{fig:depolarization} and \cref{fig:polarization}.

The depolarization---translation from \fexists to \systemf---is
straightforward: it recursively
\begin{enumerate*}
  \item [(i)] removes the shift operators $[[↑]]$ and $[[↓]]$, and
  \item [(ii)] removes the sign annotations from the type variables.
\end{enumerate*}

The polarization---translation from \systemf to \fexists---is more complex.
There exist several ways to define it since any \systemf type can
be polarized either positively or negatively.
We chose the positive translation: every term of \systemf is translated
into a value. Although this translation does not minimize the number of
inserted shifts, it is more straightforward to consistently
lift it to the level of terms.

\begin{figure}[ht]
  \begin{minipage}[t]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}
\end{figure}

The term-level translation is defined not solely for terms, but for typing
derivations (we call this translation \emph{elaboration}). 
The reason for that is that the terms of \systemf and \fexists
contain different typing information that cannot be reconstructed without the
appropriate derivation tree. For instance, the lambda expressions in \systemf do
not have the bound variable annotation, while the lambda expressions in \fexists
do; on the other hand, \fexists has richer subtyping, which cannot be expressed in
in terms of polymorphic type instantiation of \systemf.

The elaboration \fexists $\leadsto$ \systemf
annotates each judgment of an \fexists typing derivation tree
with the corresponding \systemf term. This way, 
we define elaboration for each kind of judgment:
subtyping, positive typing, negative typing, and application typing.

\paragraph{From \fexists to \systemf: Subtyping Elaboration}
The subtyping elaboration is defined in \cref{fig:subt-elaboration}.
The soundness property that is preserved by the elaboration is stated in
\cref{lemma:subt-elab-soundness}. Informally, $[[Γ ⊢ iN ≤ iM ⤳ ft]]$
guarantees that $[[ft]]$ is a \systemf term that represents 
a conversion (\ie a function) from the depolarized $[[|iN|]]$ to $[[|iM|]]$.
Symmetrically, $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ implies $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$.
\begin{itemize}
  \item Rules \ruleref{\ottdruleELNVarLabel} and \ruleref{\ottdruleELPVarLabel} 
    are trivial: if the left-hand side and the right-hand side types are the 
    same, the function converting one to another is the identity.
  \item Rules \ruleref{\ottdruleELShiftULabel} and \ruleref{\ottdruleELShiftDLabel}
    are symmetric. Since the shifts are removed during the depolarization,
    the conversion function is obtained by the recursive call to the 
    premise representing the subtyping of the required order
    (the mutual subtyping equivalence is defined as two separate judgments).
    Notice that the subtyping in the other order is still required
    so that the removal of the elaboration produces a correct
    subtyping inference tree in the original \fexists.
  \item Rule \ruleref{\ottdruleELArrowLabel} allows us to
    define elaboration between function. Since the subtyping is 
    contravariant on the argument type and covariant on the result type,
    we can use the elaboration functions acquired in the premises
    (their types are $[[|iQ| → |iP|]]$ and $[[|iN| → |iM|]]$)
    to construct the required elaboration term of type 
    $[[(|iP| → |iN|) → |iQ| → |iM|]]$.
  \item Rules 
    \ruleref{\ottdruleELForallLabel}, and \ruleref{\ottdruleELExistsLabel}
    are more involved. However, the soundness property is preserved
    by \cref{lemma:depol-subst} allowing one to distribute
    depolarization over the substitution whereby
    translating the instantiation substitution $[[σ]]$ from \fexists to \systemf.
\end{itemize}

\paragraph{From \fexists to \systemf: Typing Elaboration}
Using the subtyping elaboration, we define the elaboration for the
positive/negative typing and application typing in
\cref{fig:typing-elaboration}.
The soundness property (\cref{lemma:depolar-soundness})
guarantees that if the initial term $[[v]]$ has type $[[iP]]$ in \fexists,
then the result of the elaboration of this judgment $[[ft]]$ has
type $[[|iP|]]$ in \systemf. The negative typing has a symmetric property.
As with the subtyping elaboration, 
the rules in \cref{fig:typing-elaboration} are obtained from 
the \fexists declarative typing rules (\cref{def:declarative-typing})
by annotating each judgment with the elaboration term
so that the typing is preserved.

\begin{figure}[h]
  \ottdefnELLabeled
  \caption{Subtyping elaboration from \fexists to \systemf}
  \label{fig:subt-elaboration}
\end{figure}

\clearpage % This command will start a new page
\begin{figure}[t]
  \ottdefnTELNLabeledTwoColumns{}

  \hfill

  \ottdefnTELPAppLabeled{}
  \caption{Typing elaboration from \fexists to \systemf}
  \label{fig:typing-elaboration}
\end{figure}

\paragraph{From \systemf to \fexists}

The other direction of translation is also represented as an elaboration. The
inference rules are defined in \cref{fig:systemf-polarization}, and constitute
the rules of \systemf (\cref{fig:systemf-typing}). annotated with the
elaboration terms. The elaborated term is chosen in such a way that
the soundness property (\cref{lemma:polar-soundness}) holds: 
$[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ implies $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$.

\begin{figure}[h]
  \ottdefnELfPolLabeled{}
  \caption{Typing elaboration form from \systemf to \fexists}
  \label{fig:systemf-polarization}
\end{figure}
